---
title: "Introduction to R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Reading: Faraway (2005), Appendixes A and B

```{r}　
2+3  # R can be used as a simple calculator
exp(1)  # All the usual calculator functions are available
pnorm(1.645)  # the normal probability function
```
　
The assignment operator is "<-"

```{r}
x <- 2  # assign the value 2 to x
y <- 3  # y is 3
z <- x+y  # assign the sum to z
z  # look at the result
```

Vector and matrix valued variables are also possible.

The c(), cbind() and rbind() functions are useful for building these up.

```{r}
x <- c(1,9,7,3)  # x is a length 4 vector
x  # take a look
y <- 1:4  # ":" is useful for assigning a sequence
y  # take a look
z <- x+y # vector arithmetic works as expected
z # take a look
xm <- cbind(x,y,z) # make a matrix by binding x, y, z together column-wise
xm # take a look
t(xm) # the transpose of xm
rbind(x, y, z) # make a matrix by binding x, y, z together row-wise
```
　
Now let's look at getting some simple statistics:

```{r}
x <- rnorm(50) # generat 50 standard normal random numbers
mean(x) # the mean
var(x) # the variance
summary(x) # 5 number summary plus the mean
stem(x) # a stemplot of x
y <- rnorm(50) # more random numbers
cor(x,y) # the correlation of x and y
```

Now, let's do some nice graphics:

```{r}
hist(x) # a histogram of x
plot(x,y) # a simple scatter plot
plot(x,y,pch="+",xlab="The X axis",main="My plot") # add some more options to the plot
```

To find out more online use the "help()" command:

```{r}
help(cbind) # If you know the command you want help for
```
　
Lets see how we enter data into R.  Here is a data set for test. You can save it as a text file to a directory, say "c:/mydata/", with a file name, say "gala.data". 

```{r}
gala <- read.table("gala.data") # read the data into R
```

Or, you can change the working directory to "c:/mydata" (do this by choosing "檔案"->"變更現行目錄", and then inputting the directory name). And now you can simply read the data into R by:

```{r}
gala # take a look
```

R stores whole datasets in dataframes. It's a convenient way of keeping all the variables together.

```{r}
dim(gala) # check the dimension of the data
```

You can select a variable from the data by:

```{r}
gala$Species # pick out a particular variable
gala$Sp # unique abbreviations work
attach(gala) # makes the prepending of gala unnecessary - see next line
Species 
```

You can select a subset of the data by:

```{r}
gala[2,] # the second row
gala[,3] # the third column
gala[2,3] # the (2,3) element
gala[c(1,2,4), ] # the first, second, and fourth rows
gala[3:6,] # the third through sixth rows
gala[,-c(1,2)] # "-" indicates "everything but", this keeps all data except the first two columns
gala[gala$Sp > 300,] # select those cases whose Species are greater than 300
```
　
More details on R can be found: http://www.math.montana.edu/Rweb/Rnotes/R.html.
